---
# title: "Aufgabe 2: Point Pattern Analysis"
# author: "Christa Rohrbach"
# date: "2023-03-25"
# bibliography: references.bib
# csl: harvard-educational-review.csl
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    latex_engine: xelatex
colorlinks: yes
urlcolor: blue
classoption: landscape
header-includes:
  \usepackage{fancyhdr,lastpage,multirow,caption}
  \fancypagestyle{plain}{
    \fancyhf{}
    \fancyhead[L]{Notes on the effects of different cost raster input - SDM vs. reversed SDM}
    \fancyhead[R]{Christa Rohrbach}
    \fancyfoot[R]{\thepage\  / \pageref{LastPage}}
  }
  \pagestyle{plain}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE
)

# Exponentialschreibweise erst ab 7-stelligen Zahlen
options(scipen=6, digits=4)

library(knitr)
library(kableExtra)
library(readr)
library(cowplot)
library(ggplot2)
library(sf)
library(terra)
library(tidyterra)
library(rnaturalearth)
library(rnaturalearthdata)

# rgdal, rgeos and maptools retire by the end of 2023! use sf, stars and terra!

theme_set(theme_classic())
theme_update(
  panel.grid.major = element_line(colour = "transparent"),
  plot.title = element_text(size=6),
  #axis.title = element_text(size=10),
  #legend.title = element_text(size=8),
  axis.text = element_text(size=4),
  #axis.text.x = element_text(angle = 45, hjust=1)
  #panel.border = element_rect(color = "black", fill = NA, size = 0.1),
  axis.ticks = element_line(linewidth = 0.1),
  #axis.line.x = element_blank(),
  #axis.line.y = element_blank(),
  axis.line = element_line(linewidth = 0.1),
  axis.title = element_blank(),
  legend.title = element_text(size = 6),
  legend.text = element_text(size = 4),
  legend.key.size = unit(3, 'mm'),
  legend.box.spacing = unit(0, 'mm')
)
```

```{r settings, results='hide'}
# Input data paths and configuration
workdir <- "C:/Daten/Dokumente/UNIGIS/ArcGIS Projekte/p_distanzanalyse"
presence_gpkg <- "msculpturalis_20240106.gpkg"
presence_layer_name <- strsplit(presence_gpkg, "\\.")[[1]][1]
cost_surface_name <- "msculpturalis_sdm_clip.tif"
cost_surface_name_2 <- "msculpturalis_sdm_clip_rev.tif"
paths_gpkg <- "msculpturalis_20240106_sdm.gpkg" 
paths_gpkg_2 <- "msculpturalis_20240106_sdmrev.gpkg" 
paths_layer_name <- "msculpturalis_20240106_sdm_paths"
paths_layer_name_2 <- "msculpturalis_20240106_sdmrev_paths"
quantile <- 0.96
```

```{r data, results='hide'}
# Load data
cost_surface <- rast(file.path(workdir, cost_surface_name))
#NAflag(cost_surface) <- 0  # tbd: workaround to avoid NoData being plotted
site_bbox <- ext(cost_surface)

cost_surface_2 <- rast(file.path(workdir, cost_surface_name_2))
#cost_surface_2 <- crop(cost_surface_2, site_bbox)  # tbd: was an attempt to crop second raster to extent of first, but same CRS needed
#NAflag(cost_surface_2) <- 0  # tbd: workaround to avoid NoData being plotted

presence <- st_read(file.path(workdir, presence_gpkg), layer = presence_layer_name)

lcp <- st_read(file.path(workdir, paths_gpkg), layer = paths_layer_name)
threshold <- quantile(lcp$accumulated_cost, quantile)
lcp_sub <- lcp[lcp$accumulated_cost < threshold, ]

lcp_2 <- st_read(file.path(workdir, paths_gpkg_2), layer = paths_layer_name_2)
threshold_2 <- quantile(lcp_2$accumulated_cost, quantile)
lcp_sub_2 <- lcp_2[lcp_2$accumulated_cost < threshold_2, ]

world <- st_transform(ne_countries(scale = "medium", returnclass = "sf"), crs(cost_surface))
countries <- st_crop(world, site_bbox)
graticules <- st_graticule(x = countries)
```

# Notes on the effects of different cost raster input - SDM vs. reversed SDM

**Left side:**  
Cost raster *`r cost_surface_name`* with value range `r minmax(cost_surface)[1]`-`r minmax(cost_surface)[2]` and resulting least-cost paths.  
The cost raster is the habitat suitability raster ("consensus map") published by [Lanner et al. (2022)](https://doi.org/10.1016/j.scitotenv.2022.154246).

**Right side:**  
Cost raster *`r cost_surface_name_2`* with value range `r minmax(cost_surface_2)[1]`-`r minmax(cost_surface_2)[2]` and resulting least-cost paths.  
The cost raster is a reversed version of the habitat suitability raster ("consensus map") published by [Lanner et al. (2022)](https://doi.org/10.1016/j.scitotenv.2022.154246).

**Methodology:**  
Least-cost paths were computed using the [Python module skimage.graph (MCP_Geometric class)](https://scikit-image.org/docs/stable/api/skimage.graph.html), which applies a network-based algorithm allowing an 8-way movement from cell to cell.
Least-cost paths exceeding the accumulated cost of the `r quantile` quantile are not shown in the maps.

```{r basemap, results='hide'}
# Create basemap
basemap <- ggplot() +
  geom_sf(data = countries, fill = "#f7f7f7", color = "gray", linewidth = 0.1) +
  #coord_sf(datum = st_crs(cost_surface)) +
  geom_sf(data = graticules, color = "gray", linetype = "dashed", linewidth = 0.1)
```

```{r rasterplots, results='hide', fig.keep='all', out.width='\\textwidth', out.height='\\textheight', fig.align='top'}
# Create maps with cost rasters
cost1 <- ggplot() +
  geom_spatraster(data = cost_surface) +
  scale_fill_viridis_c(na.value = "transparent", direction = -1) +
  #geom_sf(data = countries, fill = "transparent", color = "gray", linewidth = 0.1) +
  geom_sf(data = graticules, color = "gray", linetype = "dashed", linewidth = 0.1) +
  coord_sf(expand = FALSE) +
  ggtitle(cost_surface_name)

cost2 <- ggplot() +
  geom_spatraster(data = cost_surface_2) +
  scale_fill_viridis_c(na.value = "transparent", direction = -1, name = "value (rev.)") +
  #geom_sf(data = countries, fill = "transparent", color = "gray", linewidth = 0.1) +
  geom_sf(data = graticules, color = "gray", linetype = "dashed", linewidth = 0.1) +
  coord_sf(expand = FALSE) +  # this removes white space between map and axes
  ggtitle(cost_surface_name_2)

plot_grid(cost1, cost2)
```

```{r pathplots, results='hide', fig.keep='all', out.width='\\textwidth', out.height='\\textheight', fig.align='top'}
# Create a list to store maps for each year
all_maps_list <- list()

# Loop through years 2009 to 2023
for (year in 2009:2023) {
  # Create a list to store maps for the current year
  maps_list <- list()
  
  lcp1 <- basemap +
    geom_sf(data = presence[presence$observation_year == year, ], size = 0.1, color = "red") +
    geom_sf(data = lcp_sub[lcp_sub$year == year, ], linewidth = 0.1, color = "red") +
    geom_sf(data = presence[presence$observation_year < year, ], size = 0.1, color = "#3b55ff") +
    geom_sf(data = lcp_sub[lcp_sub$year < year, ], linewidth = 0.1, color = "#3b55ff") +
    coord_sf(expand = FALSE) +
    ggtitle(paste0(year, " | ", paths_layer_name, " | cost < ", quantile, " quantile (", threshold, ")"))
  
  lcp2 <- basemap +
    geom_sf(data = presence[presence$observation_year == year, ], size = 0.1, color = "red") +
    geom_sf(data = lcp_sub_2[lcp_sub_2$year == year, ], linewidth = 0.1, color = "red") +
    geom_sf(data = presence[presence$observation_year < year, ], size = 0.1, color = "#3b55ff") +
    geom_sf(data = lcp_sub_2[lcp_sub_2$year < year, ], linewidth = 0.1, color = "#3b55ff") +
    coord_sf(expand = FALSE) +
    ggtitle(paste0(year, " | ", paths_layer_name_2, " | cost < ", quantile, " quantile (", threshold_2, ")"))
  
  # Store the maps for the current year in the list
  maps_list[[as.character(year)]] <- plot_grid(lcp1, lcp2, ncol = 2)
  
  # Store the map list for the current year in the overall list
  all_maps_list[[as.character(year)]] <- maps_list
}

# Output the maps using cowplot
plot_grid(
  all_maps_list[[as.character(2009)]][[1]],
  all_maps_list[[as.character(2010)]][[1]],
  ncol = 1
)
plot_grid(
  all_maps_list[[as.character(2011)]][[1]],
  all_maps_list[[as.character(2012)]][[1]],
  ncol = 1
)
plot_grid(
  all_maps_list[[as.character(2013)]][[1]],
  all_maps_list[[as.character(2014)]][[1]],
  ncol = 1
)
plot_grid(
  all_maps_list[[as.character(2015)]][[1]],
  all_maps_list[[as.character(2016)]][[1]],
  ncol = 1
)
plot_grid(
  all_maps_list[[as.character(2017)]][[1]],
  all_maps_list[[as.character(2018)]][[1]],
  ncol = 1
)
plot_grid(
  all_maps_list[[as.character(2019)]][[1]],
  all_maps_list[[as.character(2020)]][[1]],
  ncol = 1
)
plot_grid(
  all_maps_list[[as.character(2021)]][[1]],
  all_maps_list[[as.character(2022)]][[1]],
  ncol = 1
)
plot_grid(
  all_maps_list[[as.character(2023)]][[1]],
  ncol = 1
)
```
